\vspace{-3em} \textbf{ Mapping stellar surfaces\\ II: An interpretable Gaussian process model for light curves }

### 2021-01-29 paper2/ms.tex
finished paper 2

Moreover, even when the stellar signal is considered a nuisance, a
physically-driven variability model may be a better choice than an
effective model in some cases, particularly when the signal of interest is small
compared to the systematics. A specific example of this is in transmission
spectroscopy of transiting exoplanets, where the contribution from
unocculted spots and faculae to the spectrum
can be an order of magnitude larger than that of the planet atmosphere
\citep{Rackham2018}. In this case, failure to explicitly model the effect of starspots
can lead to spurious features in the planet spectrum.
%
A similar situation arises in extreme precision radial velocity (EPRV)
searches for planets, where the stellar signal can be orders of magnitude
larger than the planetary signal. While effective models of
variability have often been successful at disentangling the planetary
and stellar contributions \xxx{\citep{XXX}}, these models can struggle
when the (a priori unknown) orbital period of the planet
is close to an alias of the rotational period of the star \xxx{\citep{XXX}}.
In this case, a physically-driven model of variability would likely
perform better.

@@

What Equation~(\ref{eq:SigmaTilde}) allows us to do is effectively marginalize over
the unknown baseline by modeling the \emph{normalized} flux as a draw
from a Gaussian process:
%
\begin{align}
    \tilde{\mathbbb{f}\hspace{0.2em}}
    \left(P, \mathbf{u}, \pmb{\theta}_\bullet\right)
    \sim
    \mathcal{N}\left(
    \mathbf{1},
    \tilde{\pmb{\Sigma}} \left(P, \mathbf{u}, \pmb{\theta}_\bullet\right)
    \right)
    \quad.
\end{align}
%
This is appropriate as long as $z \ll 1$, for which the true
distribution of $\tilde{\mathbbb{f}\hspace{0.2em}}$ is approximately Gaussian. In practice,
we recommend employing this trick only for $z \lesssim 0.02$, for which the
error in the approximation to the covariance is less than $10^{-6}$.
In cases where the light curve variability exceeds about ten percent, we recommend
modeling the multiplicative amplitude in each light curve as a latent variable, as discussed
above.

@@

The runs so far correspond to stars with many ($n = 20$) spots, for which the
resulting light curves are smooth due to the fact that many spots are in
view at any given time. Figures~\ref{fig:calibration_hicontrast} and
\ref{fig:calibration_bigspots} show what happens when the model is applied to
stars with $n=2$ and $n=1$ spots, respectively. Despite large portions of the
light curves being flat (and therefore extremely non-stationary)
in these scenarios, the GP does surprisingly well, recovering the radii
and latitude parameters within $2-3\sigma$ in both cases.
Note that in order to preserve the same signal-to-noise ratio relative to
the other runs, we gave the spots in Figure~\ref{fig:calibration_hicontrast}
a much higher contrast ($c = 0.5$). Even though the contrast is degenerate
with the number of spots (which is unobservable), the $c$ posterior
has a much heavier tail than in the other runs. This is largely coming
from our prior, in which we assume that the spots must be dark and all have
the same contrast.
In Figure~\ref{fig:calibration_bigspots}, we compensate for the smaller number
of spots by increasing the spot radius to $r = 45^\circ \pm 5^\circ$ instead,
showing that the model can accurately model large spots, even in the
presence of some (unmodelled) scatter in their sizes.

@@

Our final trial is shown in Figure~\ref{fig:calibration_unnorm}, in which
we assume we know the true baseline of each light curve. That is, we
assume that we can measure all light curves in units of the flux we would
measure if the stars had no spots on them, and we \emph{do not}
normalize them (see \S\ref{sec:gp-norm}). In practice, this would require
knowledge of the brightness (or temperature) of the unspotted
photosphere, which is not an observable in single-band relative photometry.
This value can in principle be probed, however, in multi-band photometry
\citep[e.g.,][]{Gully2017,Guo2018}, for which this trial is
extremely relevant. We again recover the radii and latitude parameters
to within $2-3\sigma$, but most importantly, we also infer the correct
spot contrast \emph{and} the correct number of spots with fairly high
precision. In particular, knowledge of the baseline breaks the $c-N$ degeneracy.
Photometric measurements in multiple bands (even just two!)
are therefore extremely useful when inferring spot properties.
We discussed this point in \citetalias{PaperI}.

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/brightspots.pdf}
        \oscaption{brightspots}{%
            Comparison of GP samples with dark spots (top) and bright
            spots (center) alongside their corresponding light curves
            viewed at $I = 75^\circ$ (bottom). Without knowledge of
            the baseline (\S\ref{sec:gp-norm}), it is very
            difficult to differentiate between the two from stellar
            light curves.
            \label{fig:brightspots}
        }
    \end{centering}
\end{figure}

@@

The reason for this degeneracy is rooted (once again) in the
fundamental issue with relative photometry: we lack any information about
the true baseline level of the light curve.
Consider the dependence of the (unnormalized) GP covariance on the
contrast: it enters in a single place, via Equation~(\ref{eq:bigEc}),
as $c^2$, meaning dark and bright spot models have exactly the same
covariance. These models differ only in the mean of the unnormalized
process, since that is proportional to $c$ (via Equation~\ref{eq:ec}).
However, as we argue in \S\ref{sec:gp-norm}, the mean is not a direct
observable. Instead, in relative photometry, we are only sensitive to
the \emph{ratio} of the covariance to the square of the mean
(see Equation~\ref{eq:SigmaTilde}). From that equation, we can deduce
that stars with dark spots (for which
the light curve mean $\mu < 1$) will therefore have larger variance
than stars with bright spots ($\mu > 1$), leading to the slight difference
seen in the figure. However, since this is strictly a multiplicative
factor affecting the covariance, it is degenerate with the two other
properties that scale the covariance: the magnitude of the contrast and
the total number of spots.%
\footnote{There is also a small additive term in
    Equation~(\ref{eq:SigmaTilde}), but this, too, depends only on the ratio
    of entries in the covariance matrix to the mean, so it is of little help
    in breaking the degeneracy.}

@@

We can, however, skirt this degeneracy with observations in multiple bands,
which can provide limited information about the baseline level. Recently,
\citet{Morris2018} used approximately coeval \emph{Kepler} and \emph{Spitzer}
light curves of TRAPPIST-1 to argue that a bright spot model for the star
is more consistent with the data. A detailed exploration of the
effect of multi-band photometry on the degeneracies of the mapping problem
is deferred to a future paper in this series.

@@

The core idea behind the methodology presented in this paper---to compute the
Gaussian approximation to an intractable multidimensional distribution in order
to obtain a likelihood function for inference---is not new. Although the
method likely goes by different names in different fields, it is a popular
technique particularly in the field of ecological population dynamics,
where it is referred to as the synthetic likelihood \citep[SL;][]{Wood2010}
or Bayesian synthetic likelihood \citep[BSL;][]{Price2018} method. In many
ecological systems, population growth is a chaotic process; observations of the
size of a population over time can be dominated by steep spikes and drops
in the population that occur due to sudden, random environmental pressure.
While population growth can be forward modeled with ease, it is very
difficult to use forward models to constrain basic growth parameters in an
inference setting, since that requires marginalization over the extremely
nonlinear noice processes. As a way around this, \citet{Wood2010} introduced
the SL method, in which, conditioned on a set of parameters of interest
$\pmb{\theta}$, one computes the forward model $f(\pmb{\theta})$ many times
under different realizations of the noise, and adopts the sample mean
and sample covariance as the mean and covariance of the Gaussian likelihood
function $p(f \, | \, \pmb{\theta})$. \citet{Wood2010} showed that,
provided the number of forward model samples is large enough, this
``synthetic'' likelihood allows one to infer the population growth parameters
efficiently and without bias.

The method presented in this paper may be thought of as a synthetic likelihood
method in the limit of an infinite number of forward model samples. Unlike
\citet{Wood2010}, whose method determines the mean and covariance of the
distribution of $f$ conditioned on $\pmb{\theta}$ by sampling, we are able
to actually compute the mean and covariance of $f$ \emph{in closed form}.
While traditional SL methods are inherently noisy, our method employs
the \emph{exact} Gaussian approximation to the likelihood function.

Our GP is also closely related to techniques commonly employed in models
of the cosmic microwave background (CMB). In particular, it is a type of
Gaussian random field (GRF) on the sphere, which are frequently used to
model perturbations in the CMB \citep{Wandelt2012}. In general, however,
GRFs used in cosmology are isotropic: when expressed in the spherical
harmonic basis, their covariance matrix is diagonal and admits a
representation as a (one-dimensional) power spectrum. Our GP, in contrast,
is anisotropic in
the polar coordinate (i.e., the latitude) by construction.

@@

Finally, \citet{Basri2020} recently presented a large suite of forward models of light curves of
spotted stars, which they used to discuss the (complicated) dependence of
various light curve metrics on
the physical spot parameters used to generate the data. They concluded that it
is not possible to uniquely relate these metrics to the underlying starspot
configuration. While we agree this is the case for individual stars, our work
stresses that it is possible to break many of these degeneracies with
ensemble analyses.
%
\citet{Basri2020} also concluded it is not in general possible to
uniquely disentangle differential rotation from spot evolution when their
timescales are comparable; nor is it possible to confidently measure a rotation
period when the evolution timescale is very short. However, their study
relied on the effect these processes have on simple light curve metrics,
which are almost certainly not sufficient statistics of the data. Inference
that takes into account the full covariance structure of the data, while
considering large ensembles of light curves, could in principle break
this degeneracy. While we do not explicitly model differential rotation in
this paper, it will be the subject of a future paper in this series.

@@

Armed with this algorithm for computing $\pmb{\Sigma}^\mathbf{(t)}$, it only
remains for us to decide on a structure for $\mathbf{K}$. While this can
in principle be any covariance matrix constructed from a stationary kernel,
we recommend one of the common radial kernels such as the exponential
squared kernel (Equation~\ref{eq:expsq}) or the Mat\'ern-3/2 kernel
(Equation~\ref{eq:mat32}) with variance parameter $\sigma^2$ set to
unity (since the variance is already specified within $\pmb{\Sigma}$). In this
case, the temporal covariance $\mathbf{K}$ is a function of a single
parameter: the timescale of the variability, $\tau$. Similar to the other
hyperparameters of our GP, this parameter can be estimated in an inference
setting. However, a detailed investigation of the ability of our temporal
GP to accurately capture spot variability is beyond the scope of this paper,
and will be revisited in the future, along with an algorithm to explicitly
model the effects of differential rotation on the covariance structure.

@@

Consider Equation~(\ref{eq:akT}) in the Appendix, from which the rows of the
design matrix (which transforms vectors in the spherical harmonic basis to vectors
representing the flux at a point in time) are computed. Marginalization over the
inclination, which we demonstrated how to perform analytically in
Appendix~\ref{sec:inc}, entails integrating over the term
$\mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right)$, the Wigner matrix that rotates
the star by the inclination angle $I$ into the observer's frame.
Similarly, marginalization over the
rotation period would entail integration over the term
$\mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P} t_k\right)$, another
Wigner matrix that rotates the star to the correct rotational phase at
time $t = t_k$. Given the appropriate prior on the rotational angular
frequency $\frac{2\pi}{P}$, it should be possible to follow the same
procedure outlined in Appendix~\ref{sec:inc} to analytically
compute the first two moments of the distribution of light curves marginalized over the
rotation period.

@@

Given the complexity of the operations involved,
performing these marginalizations is beyond the scope of the present paper.
However, given the significant computational benefits of this marginalization,
and given the fact that the mapping problem is particularly sensitive to
the limb darkening coefficients \citepalias{PaperI}, marginalizing over
$P$ and $\mathbf{u}$ will be the subject of an upcoming paper.

@@

Finally, we would like to note that the GP developed in this paper is not
limited to modeling (rotational) light curves of stars. We derived the
covariance structure of spotted stellar surfaces in the spherical harmonic
basis, which we then linearly transformed into flux space to obtain the
light curve GP presented above. However, our GP may be used to model
\emph{any} kind of observation that
is linearly related to the spherical harmonic representation. If $\mathbf{A}$
is the linear operator that transforms the spherical harmonic representation
$\mathbf{y}$ to the data vector $\mathbf{d}$ via
$\mathbf{d} = \mathbf{A} \mathbf{y}$, then the mean and covariance of the
GP model for $\mathbf{d}$ are given by
%
\begin{align}
    \pmb{\mu}_\mathbf{d}    & = \mathbf{A} \, \pmb{\mu}_\mathbf{y}
    \\
    \pmb{\Sigma}_\mathbf{d} & = \mathbf{A} \, \pmb{\Sigma}_\mathbf{y} \, \mathbf{A}^\top
    \quad,
\end{align}
%
respectively, where $\pmb{\mu}_\mathbf{y}$ and $\pmb{\Sigma}_\mathbf{y}$
are the mean and covariance in the spherical harmonic basis derived in this paper.
In \citet{Luger2019} we showed that
occultation light curves are also linearly related to the
spherical harmonic representation of the stellar surface, so it is
straightforward to use our GP to model transits of planets across spotted
stars, either to marginalize over the spot variability or to constrain the
surface map of the star. In this context, $\mathbf{A}$ may be computed via the
\textsf{design\_matrix()} method of a \textsf{Map} instance of the \starry
package.

@@

Our goal in this section is to find closed-form solutions to the
first and second moments of the spherical harmonic representation of the
stellar surface $\mathbbb{y}$,
%
\begin{align}
    \label{eq:exp_y_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \\
    \label{eq:exp_yy_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \mathbbb{y}^\top(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \quad,
\end{align}
%
which are linearly related to the mean and the covariance of our GP
(\S\ref{sec:gp-gp}).
%
Recall that $\mathbbb{x}$ is a vector of parameters describing the exact
configuration of features on the surface of a star, and
$p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$ is its probability
density function conditioned on hyperparameters $\pmb{\theta}_\bullet$,
which describe the \emph{distribution} of the features on the surface
of one or many stars.
%
As we are specifically interested in modeling the effect of starspots
on stellar light curves, we let
%
\begin{align}
    \label{eq:x}
    \mathbbb{x} =
    \left(
    \mathbb{n} \,\,\,\,
    \mathbb{c}_0 \,\, \cdots \,\, \mathbb{c}_{\mathbb{n}-1} \,\,\,\,
    \bblambda_0 \,\, \cdots \,\, \bblambda_{\mathbb{n}-1} \,\,\,\,
    \bbphi_0 \,\, \cdots \,\, \bbphi_{\mathbb{n}-1} \,\,\,\,
    \mathbb{r}_0 \,\, \cdots \,\, \mathbb{r}_{\mathbb{n}-1}
    \right)^\top
\end{align}
%
and
%
\begin{align}
    \label{eq:RRs}
    \mathbbb{y}(\mathbbb{x}) =
    -
    \sum_{i=0}^{\mathbb{n}-1}
    \mathbb{c}_i
    \,
    \mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)
    \,
    \mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)
    \,
    \mathbf{s}(\mathbb{r}_i)
    \quad,
\end{align}
%
where $\mathbb{n}$ is the total number of spots,
$\mathbb{c}_i$ is the contrast of the $i^\mathrm{th}$ spot,
$\bblambda_i$ is its longitude, $\bbphi_i$ is its latitude,
and $\mathbb{r}_i$ is its radius.
The vector function $\mathbf{s}(\mathbb{r}_i)$
returns the spherical harmonic expansion of a negative unit brightness
circular spot of radius $\mathbb{r}_i$ at $\bblambda = \bbphi = 0$,
$\mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)$ is the Wigner matrix that rotates the
expansion about $\hat{\mathbf{x}}$ such that the spot is centered at a
latitude $\bbphi_i$, and $\mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)$ is the Wigner
matrix that then rotates the
expansion about $\hat{\mathbf{y}}$ such that the spot is centered at a
longitude $\bblambda_i$; these three functions are detailed in the sections below.
%
Equation~(\ref{eq:RRs}) thus provides a way of converting a random variable
$\mathbbb{x}$ describing the size, brightness, and position of spots to the
corresponding representation in terms of spherical harmonics.
%
Regarding this equation,
two things should be noted. First, we define $\mathbbb{y}$ relative to
a baseline of zero: i.e., a star with no spots on it will have
$\mathbf{y} = \mathbf{0}$. Second, and more importantly,
we are not interested in any specific value of
$\mathbbb{y}$; rather, we would like to know its expectation value under
the probability distribution governing the different spot properties $\mathbbb{x}$,
i.e., $p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$.
%

>>>>

Moreover, even when the stellar signal is considered a nuisance, a
physically-driven variability model may be a better choice than an
effective model in some cases, particularly when the signal of interest is small
compared to the systematics. A specific example of this is in transmission
spectroscopy of transiting exoplanets, where the contribution from
unocculted spots and faculae to the spectrum
can be an order of magnitude larger than that of the planet atmosphere
\citep{Rackham2018}. In this case, failure to explicitly model the effect of starspots
can lead to spurious features in the planet spectrum.
%
A similar situation arises in extreme precision radial velocity (EPRV)
searches for planets, where the stellar signal can be orders of magnitude
larger than the planetary signal. While effective models of
variability have often been successful at disentangling the planetary
and stellar contributions \citep[e.g.,][]{Rajpaul2015}, these models can struggle
when the (a priori unknown) orbital period of the planet
is close to an alias of the rotational period of the star \citep{Vanderburg2016,Damasso2019,Robertson2020}.
In this case, a physically-driven model of variability would likely
perform better.

@@

What Equation~(\ref{eq:SigmaTilde}) allows us to do is effectively marginalize over
the unknown normalization by modeling the \emph{normalized} flux as a draw
from a Gaussian process:
%
\begin{align}
    \tilde{\mathbbb{f}\hspace{0.2em}}
    \left(P, \mathbf{u}, \pmb{\theta}_\bullet\right)
    \sim
    \mathcal{N}\left(
    \mathbf{1},
    \tilde{\pmb{\Sigma}} \left(P, \mathbf{u}, \pmb{\theta}_\bullet\right)
    \right)
    \quad.
\end{align}
%
This is appropriate as long as $z \ll 1$, for which the true
distribution of $\tilde{\mathbbb{f}\hspace{0.2em}}$ is approximately Gaussian. In practice,
we recommend employing this trick only for $z \lesssim 0.02$, for which the
error in the approximation to the covariance is less than $10^{-6}$.
In cases where the light curve variability exceeds about ten percent, we recommend
modeling the multiplicative amplitude in each light curve as a latent variable, as discussed
above.

@@

The runs so far correspond to stars with many ($n = 20$) spots, for which the
resulting light curves are smooth due to the fact that many spots are in
view at any given time. Figures~\ref{fig:calibration_hicontrast} and
\ref{fig:calibration_bigspots} show what happens when the model is applied to
stars with $n=2$ and $n=1$ spots, respectively. Despite large portions of the
light curves being flat (and therefore extremely non-stationary)
in these scenarios, the GP does surprisingly well, recovering the radii
and latitude parameters within $2-3\sigma$ in both cases.
Note that in order to preserve the same signal-to-noise ratio relative to
the other runs, we gave the spots in Figure~\ref{fig:calibration_hicontrast}
a much higher contrast ($c = 0.5$). Even though the contrast is degenerate
with the number of spots (which is very poorly constrained), the $c$ posterior
has a much heavier tail than in the other runs.
Thus, in spite of the arguments in \citet{PaperI} about the difficulty in
constraining $c$ and $n$ from single-band photometry, it is evident that
the full covariance structure of the data encodes \emph{some} information
about the contrast and--to a much lesser extent--the number of spots.
In Figure~\ref{fig:calibration_bigspots}, we compensate for the smaller number
of spots by increasing the spot radius to $r = 45^\circ \pm 5^\circ$ instead,
showing that the model can accurately model large spots, even in the
presence of some (unmodelled) scatter in their sizes.

@@

Our final trial is shown in Figure~\ref{fig:calibration_unnorm}, in which
we assume we know the true normalization of each light curve. That is, we
assume that we can measure all light curves in units of the flux we would
measure if the stars had no spots on them, and we \emph{do not}
normalize them (see \S\ref{sec:gp-norm}). In practice, this would require
knowledge of the brightness (or temperature) of the unspotted
photosphere, which is not an observable in single-band photometry.
This value can in principle be probed, however, in multi-band photometry
\citep[e.g.,][]{Gully2017,Guo2018}, for which this trial is
extremely relevant. We again recover the radii and latitude parameters
to within $2-3\sigma$, but most importantly, we also infer the correct
spot contrast \emph{and} the correct number of spots with fairly high
precision. In particular, knowledge of the correct normalization breaks the $c-n$ degeneracy.
Photometric measurements in multiple bands (even just two!)
are therefore extremely useful when inferring spot properties.
We discussed this point in \citetalias{PaperI}.

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/brightspots.pdf}
        \oscaption{brightspots}{%
            Comparison of GP samples with dark spots (top) and bright
            spots (center) alongside their corresponding light curves
            viewed at $I = 75^\circ$ (bottom). Without knowledge of
            the correct normalization (\S\ref{sec:gp-norm}), it is very
            difficult to differentiate between the two from stellar
            light curves.
            \label{fig:brightspots}
        }
    \end{centering}
\end{figure}

@@

The reason for this degeneracy is rooted (once again) in the
fundamental issue with photometry: we lack any information about
the correct normalization of the light curve.
Consider the dependence of the (unnormalized) GP covariance on the
contrast: it enters in a single place, via Equation~(\ref{eq:bigEc}),
as $c^2$, meaning dark and bright spot models have exactly the same
covariance. These models differ only in the mean of the unnormalized
process, since that is proportional to $c$ (via Equation~\ref{eq:ec}).
However, as we argued in \citetalias{PaperI}, the mean is not a direct
observable. Instead, in single-band photometry, we are only sensitive to
the \emph{ratio} of the covariance to the square of the mean
(see Equation~\ref{eq:SigmaTilde}). From that equation, we can deduce
that stars with dark spots (for which
the light curve mean $\mu < 1$) will therefore have larger variance
than stars with bright spots ($\mu > 1$), leading to the slight difference
seen in the figure. However, since this is strictly a multiplicative
factor affecting the covariance, it is degenerate with the two other
properties that scale the covariance: the magnitude of the contrast and
the total number of spots.%
\footnote{There is also a small additive term in
    Equation~(\ref{eq:SigmaTilde}), but this, too, depends only on the ratio
    of entries in the covariance matrix to the mean, so it is of little help
    in breaking the degeneracy.}

@@

We can, however, skirt this degeneracy with observations in multiple bands,
which can provide limited information about the correct normalization. Recently,
\citet{Morris2018} used approximately coeval \emph{Kepler} and \emph{Spitzer}
light curves of TRAPPIST-1 to argue that a bright spot model for the star
is more consistent with the data. A detailed exploration of the
effect of multi-band photometry on the degeneracies of the mapping problem
is deferred to a future paper in this series.

@@

The core idea behind the methodology presented in this paper---to compute the
Gaussian approximation to an intractable multidimensional distribution in order
to obtain a likelihood function for inference---is not new. Although the
method likely goes by different names in different fields, it is a popular
technique particularly in the field of ecological population dynamics,
where it is referred to as the synthetic likelihood \citep[SL;][]{Wood2010}
or Bayesian synthetic likelihood \citep[BSL;][]{Price2018} method. In many
ecological systems, population growth is a chaotic process; observations of the
size of a population over time can be dominated by steep spikes and drops
in the population that occur due to sudden, random environmental pressure.
While population growth can be forward modeled with ease, it is very
difficult to use forward models to constrain basic growth parameters in an
inference setting, since that requires marginalization over the extremely
nonlinear noice processes. As a way around this, \citet{Wood2010} introduced
the SL method, in which, conditioned on a set of parameters of interest
$\pmb{\theta}$, one computes the forward model $f(\pmb{\theta})$ many times
under different realizations of the noise, and adopts the sample mean
and sample covariance (usually of a summary statistic of the data)
as the mean and covariance of the Gaussian likelihood
function $p(f \, | \, \pmb{\theta})$. \citet{Wood2010} showed that,
provided the number of forward model samples is large enough, this
``synthetic'' likelihood allows one to infer the population growth parameters
efficiently and without bias.

The method presented in this paper may be thought of as a synthetic likelihood
method in the limit of an infinite number of forward model samples. Unlike
\citet{Wood2010}, whose method determines the mean and covariance of the
distribution of some function of $f$ conditioned on $\pmb{\theta}$ by sampling, we are able
to actually compute the mean and covariance of $f$ directly \emph{in closed form}.
While traditional SL methods are inherently noisy, our method employs
the \emph{exact} Gaussian approximation to the likelihood function.

Our GP is also closely related to techniques commonly employed in models
of the cosmic microwave background (CMB). In particular, it is a type of
Gaussian random field (GRF) on the sphere, which is frequently used to
model perturbations in the CMB \citep{Wandelt2012}. In general, however,
GRFs used in cosmology are isotropic: when expressed in the spherical
harmonic basis, their covariance matrix is diagonal and admits a
representation as a (one-dimensional) power spectrum. Our GP, in contrast,
is anisotropic in
the polar coordinate (i.e., the latitude) by construction.

@@

Finally, \citet{Basri2020} recently presented a large suite of forward models of light curves of
spotted stars, which they used to discuss the (complicated) dependence of
various light curve metrics on
the physical spot parameters used to generate the data. They concluded that it
is not possible to uniquely relate these metrics to the underlying starspot
configuration. While we agree this is the case for individual stars, our work
stresses that it is possible to circumvent many of these degeneracies with
ensemble analyses.
%
\citet{Basri2020} also concluded it is not in general possible to
uniquely disentangle differential rotation from spot evolution when their
timescales are comparable; nor is it possible to confidently measure a rotation
period when the evolution timescale is very short. However, their study
relied on the effect these processes have on simple light curve metrics,
which are almost certainly not sufficient statistics of the data. Inference
that takes into account the full covariance structure of the data, while
considering large ensembles of light curves, could in principle break
this degeneracy. While we do not explicitly model differential rotation in
this paper, it will be the subject of a future paper in this series.

@@

Armed with this algorithm for computing $\pmb{\Sigma}^\mathbf{(t)}$, it only
remains for us to decide on a structure for $\mathbf{K}$. While this can
in principle be any covariance matrix constructed from a stationary kernel,
we recommend one of the common radial kernels such as the exponential
squared kernel
%
\begin{align}
    \label{eq:expsq}
    k_\mathrm{E^2}(\Delta t) & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
\end{align}
%
or the Mat\'ern-3/2 kernel
\begin{align}
    \label{eq:mat32}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
\end{align}
%
with variance parameter $\sigma^2$ set to
unity (since the variance is already specified within $\pmb{\Sigma}$). In this
case, the temporal covariance $\mathbf{K}$ is a function of a single
parameter: the timescale of the variability, $\tau$. Similar to the other
hyperparameters of our GP, this parameter can be estimated in an inference
setting. However, a detailed investigation of the ability of our temporal
GP to accurately capture spot variability is beyond the scope of this paper,
and will be revisited in the future, along with an algorithm to explicitly
model the effects of differential rotation on the covariance structure.

@@

Consider Equation~(\ref{eq:akT}) in the Appendix, from which the rows of the
design matrix (which transforms vectors in the spherical harmonic basis to vectors
representing the flux at a point in time) are computed. Marginalization over the
inclination, which we demonstrated how to perform analytically in
Appendix~\ref{sec:inc}, entails integrating over the term
$\mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right)$, the Wigner matrix that rotates
the star by the inclination angle $I$ into the observer's frame.
Similarly, marginalization over the
rotation period would entail integration over the term
$\mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P} t_k\right)$, another
Wigner matrix that rotates the star to the correct rotational phase at
time $t = t_k$. Given an appropriate prior on the rotational angular
frequency $\frac{2\pi}{P}$, it should be possible to follow the same
procedure outlined in Appendix~\ref{sec:inc} to analytically
compute the first two moments of the distribution of light curves marginalized over the
rotation period.

@@

Given the complexity of the operations involved,
performing these marginalizations is beyond the scope of the present paper.
However, given the significant computational benefits of this marginalization,
as well as the fact that the mapping problem is particularly sensitive to
the limb darkening coefficients \citepalias{PaperI}, marginalizing over
$P$ and $\mathbf{u}$ will be the subject of an upcoming paper.

@@

Finally, we would like to note that the GP developed in this paper is not
limited to modeling rotational light curves of stars. We derived the
covariance structure of spotted stellar surfaces in the spherical harmonic
basis, which we then linearly transformed into flux space to obtain the
light curve GP presented above. However, our GP may be used to model
\emph{any} kind of observation that
is linearly related to the spherical harmonic representation. If $\mathbf{A}$
is the linear operator that transforms the spherical harmonic representation
$\mathbf{y}$ to the data vector $\mathbf{d}$ via
$\mathbf{d} = \mathbf{A} \mathbf{y}$, then the mean and covariance of the
GP model for $\mathbf{d}$ are given by
%
\begin{align}
    \pmb{\mu}_\mathbf{d}    & = \mathbf{A} \, \pmb{\mu}_\mathbf{y}
    \\
    \pmb{\Sigma}_\mathbf{d} & = \mathbf{A} \, \pmb{\Sigma}_\mathbf{y} \, \mathbf{A}^\top
    \quad,
\end{align}
%
respectively, where $\pmb{\mu}_\mathbf{y}$ and $\pmb{\Sigma}_\mathbf{y}$
are the mean and covariance in the spherical harmonic basis derived in this paper.
In \citet{Luger2019} we showed that
occultation light curves are also linearly related to the
spherical harmonic representation of the stellar surface, so it is
straightforward to use our GP to model transits of planets across spotted
stars, either to marginalize over the spot variability or to constrain the
surface map of the star. In this context, $\mathbf{A}$ may be computed via the
\textsf{design\_matrix()} method of a \textsf{Map} instance of the \starry
package.

@@

Our goal in this section is to find closed-form solutions to the
first and second moments of the spherical harmonic representation of the
stellar surface $\mathbbb{y}$,
%
\begin{align}
    \label{eq:exp_y_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \\
    \label{eq:exp_yy_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \mathbbb{y}^\top(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \quad,
\end{align}
%
which are linearly related to the mean and the covariance of our GP
(\S\ref{sec:gp-gp}).
%
Recall that $\mathbbb{x}$ is a vector of parameters describing the exact
configuration of features on the surface of a star, and
$p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$ is its probability
density function conditioned on hyperparameters $\pmb{\theta}_\bullet$,
which describe the \emph{distribution} of the features on the surface
of one or many stars.
%
As we are specifically interested in modeling the effect of starspots
on stellar light curves, we let
%
\begin{align}
    \label{eq:x}
    \mathbbb{x} =
    \left(
    \mathbb{n} \,\,\,\,
    \mathbb{c}_0 \,\, \cdots \,\, \mathbb{c}_{\mathbb{n}-1} \,\,\,\,
    \bblambda_0 \,\, \cdots \,\, \bblambda_{\mathbb{n}-1} \,\,\,\,
    \bbphi_0 \,\, \cdots \,\, \bbphi_{\mathbb{n}-1} \,\,\,\,
    \mathbb{r}_0 \,\, \cdots \,\, \mathbb{r}_{\mathbb{n}-1}
    \right)^\top
\end{align}
%
and
%
\begin{align}
    \label{eq:RRs}
    \mathbbb{y}(\mathbbb{x}) =
    -
    \sum_{i=0}^{\mathbb{n}-1}
    \mathbb{c}_i
    \,
    \mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)
    \,
    \mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)
    \,
    \mathbf{s}(\mathbb{r}_i)
    \quad,
\end{align}
%
where $\mathbb{n}$ is the total number of spots,
$\mathbb{c}_i$ is the contrast of the $i^\mathrm{th}$ spot,
$\bblambda_i$ is its longitude, $\bbphi_i$ is its latitude,
and $\mathbb{r}_i$ is its radius.
The vector function $\mathbf{s}(\mathbb{r}_i)$
returns the spherical harmonic expansion of a negative unit brightness
circular spot of radius $\mathbb{r}_i$ at $\bblambda = \bbphi = 0$,
$\mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)$ is the Wigner matrix that rotates the
expansion about $\hat{\mathbf{x}}$ such that the spot is centered at a
latitude $\bbphi_i$, and $\mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)$ is the Wigner
matrix that then rotates the
expansion about $\hat{\mathbf{y}}$ such that the spot is centered at a
longitude $\bblambda_i$; these three functions are detailed in the sections below.
%
Equation~(\ref{eq:RRs}) thus provides a way of converting a random variable
$\mathbbb{x}$ describing the size, brightness, and position of spots to the
corresponding representation in terms of spherical harmonics.
%
Regarding this equation,
two things should be noted. First, we define $\mathbbb{y}$ relative to
a baseline of zero: i.e., a star with no spots on it will have
$\mathbf{y} = \mathbf{0}$ (which is why we add unity in the expression for the flux
in Equation~\ref{eq:fAy}). Second, and more importantly,
we are not interested in any specific value of
$\mathbbb{y}$; rather, we would like to know its expectation value under
the probability distribution governing the different spot properties $\mathbbb{x}$,
i.e., $p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$.
%
